Theoretical prediction of disrupted Min oscillation in flattened \emph{Escherichia coli}

### 2015-03-18 paper/plos.tex
a few edits in response to Jeff

However, Mannik \emph{et al.} have recently shown that there are
limitations to this robust capability to
oscillate~\cite{mannik2012robustness,mannik2009bacterial,mannik2010bacteria}. They
have experimentally forced \emph{E. coli} cells into microfabricated
channels of thickness less than that of the natural diameter of the
cells, in which they area able to grow and divide. Upon entering the
channels, the cells undergo a mechanical deformation in which they
both widen and lengthen within the plane of the channel.  This
deformation results in very wide (they can reach widths of over
$5\micron$) flattened cells that when viewed from the top down have
irregular and asymmetric shapes.  While these cells are still able to
divide into surprisingly equal volumes, the MinD oscillations in these
cells are irregular, both temporally and spatially. Seen with
fluorescent microscopy, the MinD maximize in multiple locations within
the cell, in a seemingly random sequence. These experiments allow for
an opportunity to test MinD simulation models against more extreme
cases than have been seen thus far.

A number of models of the MinD protein system have been developed that
accurately describe its basic oscillatory nature.
%
Early models involved free proteins that affect each others' rates of
diffusion and membrane attachment, but do not combine into compound
states~\cite{meinhardt2001pattern}.  In 2003 Huang improved upon this
approach with a simple and very successful simulation model based on
MinD-MinE combination, ATPase hydrolysis, and MinD membrane attachment
that exhibits accurate MinD oscillations in cylindrical
cells~\cite{huang2003dynamic}. In this model cytoplasmic MinD is
recruited to the membrane by MinD that is already clustered there
(following observed non-linear attachment of MinD on the cell
membrane~\cite{hu2002dynamic,shih2002division}), and is stationary
once attached.
%
\blue{(5)} Several models
\cite{fange2006noise,bonny2013membrane,halatek2012highly} modify
Huang's original model to more accurately treat experimental findings
published after his 2003 paper
\cite{meacci2006mobility,loose2011min}. These models introduce slow
diffusion along the membrane (two to three orders of magnitude slower
than that within the cytoplasm), revise the process of MinD
recruitment to the membrane, and change some of the constants used in
the model. The Bonny 2013 model assumes an instantaneous nucleotide
exchange rate, and allows MinE to remain independently attached to the
membrane after dissasociating with MinD~\cite{bonny2013membrane}.
%

@@

\blue{(5a),(7)} Both Fig.~\ref{image-p} and its associated movies (see
supplementary material), show that the stochastic model exhibits a
``starry night'' effect characterized by spatially fixed points of
protein density build up, as recruitment leads to clusters of MinD
forming on the wall and then subsequently dissipating.  We attribute
this effect to the fact that the Huang model does not include
diffusion along the membrane, which has been experimentally verified
yet found to be two orders of magnitude less than diffusion in the
cytoplasm~\cite{meacci2005min}. Even ignoring this effect, however, it
is clear from Fig.~\ref{image-p} that the stochastic simulation
results in a protein distribution that is less spatially regular than
the deterministic model predicts.  This naturally leads to
irregularity in the temporal periodicity of the system, which we
quantify by examining the temporal correlation function of the total
MinD found in two opposite polar regions.
%
This correlation function, which is displayed in Fig.~\ref{corr-pill},
is given by
\begin{equation}
  C(\tau) \propto \int
  (N_{\textit{top}}(t) - \bar N_{\textit{top}})
  (N_{\textit{bottom}}(t+\tau) - \bar N_{\textit{bottom}})dt
\end{equation}
where $N_{\textit{top}}(t)$ and $N_{\textit{bottom}}(t)$ are the total
MinD proteins in the top and bottom thirds of a cell.  These plots
help us in studying the periods and the regularity and stability of
oscillations.  These curves are well fit with a simple decoherence
model, given by the equation
\begin{equation}
  C(\tau) = -\cos\left(\frac{2\pi\tau}{T}\right) e^{-\frac{\tau}{\tau_c}}
\end{equation}
where $T$ is the period and $\tau_c$ is the coherence time.  In the
case of the stochastic model the pill exhibits coherence times of
about 8 periods and in the case of the deterministic model oscillate
indefinitely with complete coherence, indicating that its behavior is
perfectly periodic.

@@

In every case, including the wild-type pill shape, we see irregularity
in the location of the maxima when using the stochastic model.  The
deterministic model shows uniformly bipolar oscillation. \blue{(2)}
It's worth pointing out that this is a significant result, since for
irregular shapes it would be reasonable to hypothesize a chaotic
system, which would be shown by irregular determinisitic data.  From
these results and how they compare with experiment, we conclude that
the deterministic model is inadequate to explain experimental
observations of the locations of density maxima of the MinD protein.

>>>>

However, Mannik \emph{et al.} have recently shown that there are
limitations to this robust capability to
oscillate~\cite{mannik2012robustness,mannik2009bacterial,mannik2010bacteria}. They
have experimentally forced \emph{E. coli} cells into microfabricated
channels of thickness less than that of the natural diameter of the
cells, in which they area able to grow and divide. Upon entering the
channels, the cells undergo a mechanical deformation in which they
both widen and lengthen within the plane of the channel.  This
deformation results in very wide (they can reach widths of over
$5\micron$) flattened cells that when viewed from the top down have
irregular and asymmetric shapes.  While these cells are still able to
divide into surprisingly equal volumes, the MinD oscillations in these
cells are irregular, both temporally and spatially. Seen with
fluorescent microscopy, the MinD maximize in multiple locations within
the cell, in a seemingly random sequence. These experiments allow for
an opportunity to test MinD simulation models against more extreme
cases than have been seen thus far.  \fixme{\blue{(2)} An additional
  interesting question is whether the irregularity observed by Mannik
  \emph{et al.} indicates that the MinD system can become chaotic (as
  in e.g. Ref.~\citenum{kuramoto1978diffusion}), or if this
irregularity is the result of a stochastic but nonchaotic process.}
\fixme{Jeff, please skim this chaos citation, which I cannot access
  from here.}

A number of models of the MinD protein system have been developed that
accurately describe its basic oscillatory nature.
%
Early models involved free proteins that affect each others' rates of
diffusion and membrane attachment, but do not combine into compound
states~\cite{meinhardt2001pattern}.  In 2003 Huang improved upon this
approach with a simple and very successful simulation model based on
MinD-MinE combination, ATPase hydrolysis, and MinD membrane attachment
that exhibits accurate MinD oscillations in cylindrical
cells~\cite{huang2003dynamic}. In this model cytoplasmic MinD is
recruited to the membrane by MinD that is already clustered there
(following observed non-linear attachment of MinD on the cell
membrane~\cite{hu2002dynamic,shih2002division}), and is stationary
once attached.
%
\blue{(5)} Several models
\cite{fange2006noise,bonny2013membrane,halatek2012highly} modify
Huang's original model to more accurately treat experimental findings
published after his 2003 paper
\cite{meacci2006mobility,loose2011min}. These models introduce slow
diffusion along the membrane (two to three orders of magnitude slower
than that within the cytoplasm), revise the process of MinD
recruitment to the membrane, and change some of the constants used in
the model.  In addition, Bonny \emph{et al.} assume an instantaneous
nucleotide exchange rate, and allows MinE to remain independently
attached to the membrane after dissasociating with
MinD~\cite{bonny2013membrane}.
%

@@

\blue{(5a),(7)} Both Fig.~\ref{image-p} and its associated movies (see
supplementary material), show that the stochastic model exhibits a
``starry night'' effect characterized by spatially fixed points of
protein density build up, as recruitment leads to clusters of MinD
forming on the wall and then subsequently dissipating.  This results
from the lack of inclusion of diffusion along the membrane in Huang's
model, which would spread out these recruitment clusters.  Experiments
have confirmed that diffusion along the membrane does occur, albeit
with a rate two orders of magnitude lower than that in the
cytoplasm~\cite{meacci2005min}.  \fixme{I would really like to see an
  estimate of the diffusivity width here.  A ``star'' from the starry
  night effect lasts about 10 seconds, so we could just figure out how
  far the protein would diffuse in 5 or 10 seconds.  If it comes out
  to close to 1 micron, we can then say that we do not expect this
  starry night effect to be observed in wild-type cells.}  \fixme{I'm
  not sure what the following sentence means.}  Even ignoring this
effect, however, it is clear from Fig.~\ref{image-p} that the
stochastic simulation results in a protein distribution that is less
spatially regular than the deterministic model predicts.  This
naturally leads to irregularity in the temporal periodicity of the
system, which we quantify by examining the temporal correlation
function of the total MinD found in two opposite polar regions.
%
This correlation function, which is displayed in Fig.~\ref{corr-pill},
is given by
\begin{equation}
  C(\tau) \propto \int
  (N_{\textit{top}}(t) - \bar N_{\textit{top}})
  (N_{\textit{bottom}}(t+\tau) - \bar N_{\textit{bottom}})dt
\end{equation}
where $N_{\textit{top}}(t)$ and $N_{\textit{bottom}}(t)$ are the total
MinD proteins in the top and bottom thirds of a cell.  These plots
help us in studying the periods and the regularity and stability of
oscillations.  These curves are well fit with a simple decoherence
model, given by the equation
\begin{equation}
  C(\tau) = -\cos\left(\frac{2\pi\tau}{T}\right) e^{-\frac{\tau}{\tau_c}}
\end{equation}
where $T$ is the period and $\tau_c$ is the coherence time.  In the
case of the stochastic model the pill exhibits coherence times of
about 8 periods and in the case of the deterministic model oscillate
indefinitely with complete coherence, indicating that its behavior is
perfectly periodic.

@@

In every case, including the wild-type pill shape, we see irregularity
in the location of the maxima when using the stochastic model.  The
deterministic model shows uniformly bipolar oscillation.  \blue{(2)}
\fixme{We conclude that the model is \emph{not} chaotic (which would
  show irregularity in the deterministic computation), but rather that
  the irregularity results purely from stochastic processes.}  From
these results and how they compare with experiment, we conclude that
the deterministic model is inadequate to explain experimental
observations of the locations of density maxima of the MinD protein.
